The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-11 rational.tex
sat

\paragraph*{Trace.}
Every compact closed categoy admits a trace. In the context of {{langRev}}
with {{eta+}}/{{eps+}}, we get the following construction:

@@

We illustrate the expressive power of first-class constraints represented by
fractional types

%%%%%%%%%%%%%%%%%%%
\subsection{Constructions with Booleans}

Before we start, let us recall a construction of conditionals from
prior work \cite{infeffects}.  Given any combinator {{c : b <-> b}} we
can construct a combinator called {{if_c : bool*b <->bool*b}} in terms
of {{c}}, where {{if_c}} behaves like a one-armed
$\mathit{if}$-expression. If the supplied boolean is {{true}} then the
combinator {{c}} is used to transform the value of type~{{b}}. If the
boolean is {{false}}, then the value of type {{b}} remains
unchanged. We can write down the combinator for {{if_c}} in terms of
{{c}} as {{ dist (;) ((id (*) c) (+) id) (;) factor }}.

% Let us look at the combinator pictorially as if it were a circuit and values
% are like particles that flow through this circuit.

The diagram below shows the input value of type {{(1+1)*b}} processed
by the distribute operator {{dist}}, which converts it into a value of
type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
case when the boolean is {{true}} (i.e. the value was {{left ()}}),
the combinator~{{c}} is applied to the value of type~{{b}}. The right
branch which corresponds to the boolean being {{false}} passes along
the value of type {{b}} unchanged.

@@

\emph{Boolean Operation based on {{if_c}}.}
The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
negates its second argument if the first argument is {{true}}. This
gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
equivalent construction that is useful is {{else_{not} }} where we
negate the second argument only is the first is {{false}}.

Similarly, we can iterate the construction of {{if_c}} to check
several bits. The gate {{if_{cnot} }}, which we may also write as
{{if^2_{not} }}, checks two booleans and negates the result wire only
if they are both {{true}}. The gate {{if^2_{not} }} is well known as
the universal reversible primitive gate called the `Toffoli gate'
\cite{Toffoli:1980}.  We can generalize this construction to
     {{if^n_{not} }} which checks {{n}} bits and negates the result
     wire only if they are all {{true}}.

\emph{Cloning Bits.} The gate {{else_{not} }} has the interesting
property that as long as the second argument is fixed to be {{true}},
output of the second wire is always the same as teh value of the first
wire. In other words {{else_{not} }} can be thought of as cloning the
value of the first wire. A circuit of {{n}} parallel {{else_{not} }}
gates can hence clone {{n}} bits. They also consume {{n}} {{true}}
inputs in the process. Let us call this construction 
{{clone^n_{bool} }}.

DIAGRAM clone

DIAGRAM clone n bits

\subsection{Construction of the Solver}
The key insight underlying the construction comes from the fact that
we can build \emph{annihilation circuits}.  Consider the circuit below
that has no observable output, i.e., for all inputs it is
\emph{A Total Annihilation Circuit.}  Consider the circuit below that
has no observable output, i.e., for all inputs it is
annihilating. This happens because it constructs a boolean and its
dual, negates one of them and attempts to collapse them.

\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/not_trace.pdf}
}
\end{center}

\emph{Controlled Annihilation.}  We leverage the basic idea of the
above circuit, to collapse all values that are not solutions to the
specified isomorphic constraint {{f}}.  The contraint {{f}} checks the
inputs to see if they match a constraint -- in the case of the SAT
solver, {{f}} would be the circuit expressing the boolean expression
to be satisfied.  While the output of {{f}} may contain many
components, its must contain a boolean indicating if the input satisfy
{{f}}. The operation {{f}} acts on some inputs and its adjoint,
{{f^{dagger} }}, reconstructs the inputs from the outputs.  The
`trick' is the following: the special output wire is used to `control'
the negation of a boolean `control wire'.  If the inputs do not
satisfy {{f}}, the control wire is negated.

@@

If this construction is traced using {{trace*}}, all the values that
do not satisfy {{f}} get annihilated because the control wire acts
like the closed-loop `not' in the previous construction.

\emph{Controlling the heap.}  The final detail we have to be concerned
with is the handling of the heap and garbage of {{f}}. Any boolean
expression {{bool^n -> bool}} can be compiled into the isomorphism
{{f:bool^h*bool^n<->bool^g*bool}} where the extra bits {{bool^h}} and
{{bool^g}} are the heap and garbage. Constructing such an isomorphism
has been detailed before \cite{Toffoli:1980,infeffects}.  However the
computation of {{f}} is meaningful {{iff}} the heap {{bool^h}} is
initialized appropriately. We can ensure that the heap has the
appropriate initial value by checking the heap and negating a second
control wire, if the values do not match, i.e., the dotted part in the
diagram below and is {{if^h_{not} }} followed by a {{not}} operator on
the control wire.
control wire, if the values do not match, i.e., the dotted part in
the diagram below and is {{if^h_{not} }} followed by a {{not}} on the
`heap control wire'.

@@

\emph{Cloning the Solution.}.  Let us call the above construction
which maps inputs, heap and control wires to input, heap and control
wires as {{sat_f}}. The SAT-solver is is completed by tracing the
{{sat_f}} and cloning the `inputs' using {{clone^n_{bool} }}.

@@

When the solver is fed inputs initialized to {{true}}, the {{clone}}
will clone any inputs to {{sat_f}} that satisfies the constraints
imposed by {{f}} and the constraints on the {{heap}}. In the case of
unique-SAT the solver will produce exactly 1 or 0 solutions. In the
case of general SAT the solver will produce solutions as determined by
the top-level (Sec. \ref{sec:rat}).

>>>>

\paragraph*{Trace.}
Every compact closed categoy admits a trace. In the context of {{langRev}}
with {{eta+}}/{{eps+}}, we get the following definition. 
Given {{f : a+c <-> b+c }}, define {{trace+ f : a <-> b}} as:

{{ trace+ f = zeroi (;) (id (+) eta+) (;) (f (+) id) (;) (id (+) eps+) (;) zeroe }}

For readability, we have omitted some the commutativity and
associativity shuffling.

@@

We illustrate the expressive power of first-class constraints represented by
fractional types. To understand the intuition, recall the definition of
{{trace}} for the multiplicative fragment of {{langRevEE}}. 
Given {{f : a*c <-> b*c }}, we have {{trace* f : a <-> b}}:

{{ trace* f = uniti (;) (id (*) eta*) (;) (f (*) id) (;) (id (*) eps*) (;) unite }}

\noindent This circuit uses {{eta*}} to generate all possible {{c}}-values
together with an associated {{(1/c)}}-constraint. It then applies {{f}} to the
pair {{(a,c)}}. The function {{f}} must produce an output {{(b,c')}} for each
such input. If the input {{c}} and the output {{c'}} are the same they can be
annihilated by {{eps*}}; otherwise the execution gets stuck and this
particular choice of {{c}} is pruned.

This construction can be used to solve a large class of constraint
satisfaction problems. We illustrate the main ideas with the implementation
of a SAT-solver. Although the main idea is the use of {{trace*}}, the
construction is involved so we proceed in small steps, reviewing some of the
necessary constructions presented in our earlier paper~\cite{infeffects}.

%%%%%%%%%%%%%%%%%%%
\subsection{Booleans and Conditionals}

Given any combinator {{c : b <-> b}} we can construct a combinator called
{{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
the combinator {{c}} is used to transform the value of type~{{b}}. If the
boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
write down the combinator for {{if_c}} in terms of {{c}} as 
{{ dist (;) ((id (*) c) (+) id) (;) factor }}.

\noindent The diagram below shows the input value of type {{(1+1)*b}}
processed by the distribute operator {{dist}}, which converts it into a value
of type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
case when the boolean is {{true}} (i.e. the value was {{left ()}}), the
combinator~{{c}} is applied to the value of type~{{b}}. The right branch
which corresponds to the boolean being {{false}} passes along the value of
type {{b}} unchanged.

@@

The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
negates its second argument if the first argument is {{true}}. This
gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
equivalent construction that is useful is {{else_{not} }} where we
negate the second argument only is the first is {{false}}.

Similarly, we can iterate the construction of {{if_c}} to check several
bits. The gate {{if_{cnot} }}, which we may also write as {{if^2_{not} }},
checks two booleans and negates the result wire only if they are both
{{true}}. The gate {{if^2_{not} }} is well known as the Toffoli gate and is a
universal reversible gate. We can generalize this construction to
{{if^n_{not} }} which checks {{n}} bits and negates the result wire only if
they are all {{true}}.

%%%%%%%%%%%%%%%%%%
\subsection{Cloning}

Although cloning is generally not allowed in reversible languages, it is
possible at the cost of having additional constant inputs. For example,
consider the gate {{else_{not} }}. Generally, the gate gates maps
{{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
on the cases in which the second input is {{true}}, we get that the gate maps
{{(false,true)}} to {{(false,false)}} and {{(true,true)}} to {{(true,true)}},
i.e., the gate clones the first input. A circuit of {{n}} parallel
{{else_{not} }} gates can hence clone {{n}} bits.  They also consume {{n}}
{{true}} inputs in the process.  Let us call this construction
{{clone^n_{bool} }}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Construction of the Solver}

The key insight underlying the construction comes from the fact that we can
build \emph{annihilation circuits} such as the one below:

\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/not_trace.pdf}
}
\end{center}
The circuit constructs a boolean {{b}} and its dual {{1/b}}, negates one of
them and attempts to satisfy the constraint that they are equal which
evidently fails.

With a little work, we can modify this circuit to only annihilate values that
fail to satisfy the constraints reprsented by a SAT-instance {{f}}. In more
detail, an instance of SAT is a function/circuit {{f}} that given some
boolean inputs returns {{true}} or {{false}} which we interpret as whether
the inputs satisfy the constraints imposed by the structure of {{f}}. Because
we are in a reversible world, our instance of SAT must be expressed as an
isomorphism: this is easily achieved as shown in Sec.~\ref{sub:f}
below. Assuming that {{f}} is expressed as an isomorphism, we have enough
information to reconstruct the input from the output. This can be done by
using the adjoint of {{f}}, denoted {{f^{dagger}}}. At this point we have,
the top half of the construction below:

@@

To summarize, the top half of the circuit is the identity function except
that we have also managed to produce a boolean wire labeled
\textsf{satisfied?} that tells us if the inputs satisfy the desired
constraints. We can take this boolean value and use it to decide whether to
negate the control wire or not. Thus, the circuit achieves the following
goal: if the inputs do not satisfy {{f}}, the control wire is negated.  We
can now use {{trace*}} to annihilate all these bad values because the control
wire acts like the closed-loop {{not}} in the previous construction.

%%%%%%%%%
\subsection{Final Details}
\label{sub:f}

Any boolean expression {{f : bool^n -> bool}} can be compiled into the
isomorphism {{iso : bool^h*bool^n<->bool^g*bool}} where the extra bits
{{bool^h}} and {{bool^g}} are considered as heap and garbage. Constructing
such an isomorphism has been detailed
before~\cite{Toffoli:1980,infeffects}. The important relation to note is that
applying {{iso}} to some special heap values and an input \textit{bs}
produces some bits that can be ignored and the same output that {{f}} would
have produced on {{bs}}. We can ensure that the heap has the appropriate
initial values by checking the heap and negating a second control wire, if
the values do not match, i.e., the dotted part in the diagram below.

@@

Let us call the above construction which maps inputs, heap, and control wires
to input, heap, and control wires as {{sat_f}}. The SAT-solver is is
completed by tracing the {{sat_f}} and cloning the inputs using
{{clone^n_{bool} }}.

@@

When the solver is fed inputs initialized to {{true}}, the {{clone}} will
clone any inputs to {{sat_f}} that satisfy the constraints imposed by {{f}}
and the constraints on the {{heap}}. In the case of unique-SAT the solver
will produce exactly 0 or 1 solutions. In the case of general SAT, the solver
will produce solutions as determined by the semantics of the top-level
interaction as discussed in Sec. \ref{sec:rat}.
